---
title: "Meat"
author: "Lizzie Kriz"
date: "2026-02-26"
output: 
  ioslides_presentation:
    incremental: true
    transition: faster
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = FALSE)
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
library(knitr)
library(tidyr)
library(tinytable)

#load in data
#meat + gdp (mgdp)- def filter this
url <- 'https://ourworldindata.org/grapher/meat-consumption-vs-gdp-per-capita.csv?v=1&csvType=full&useColumnShortNames=true'
meat_gdp <- read_csv(url)
mgdp<- meat_gdp %>% rename(Region = owid_region, meat= meat__total__00002943__food_available_for_consumption__0645pc__kilograms_per_year_per_capita,
gdp_per_capita = ny_gdp_pcap_pp_kd) %>% filter(year %in% c(1990:2025)) %>% 
  mutate(gdp_per_capita = ifelse(is.na(gdp_per_capita), 0, gdp_per_capita))

carnivores<-c('United States', 'Argentina', 'Australia')
high_risk<-c('Yemen', 'Afghanistan', 'Haiti')
```


```{r, consumption}
#meat consumed per capita (mpc)
url <- 'https://ourworldindata.org/grapher/per-capita-meat-consumption-by-type-kilograms-per-year.csv?v=1&csvType=full&useColumnShortNames=true'
OGmeat_percap <- read_csv(url)
meat_percap<- OGmeat_percap %>% filter(entity %in% c('United States', 'Argentina', 'Australia', 'Haiti', 'Afghanistan', 'Yemen'))
mpc <-
  meat_percap %>%
  rename(other = meat__other__00002735__food_available_for_consumption__0645pc__kilograms_per_year_per_capita,
        sheep_goat = meat__sheep_and_goat__00002732__food_available_for_consumption__0645pc__kilograms_per_year_per_capita, beef = meat__beef_and_buffalo__00002731__food_available_for_consumption__0645pc__kilograms_per_year_per_capita,
        pig = meat__pig__00002733__food_available_for_consumption__0645pc__kilograms_per_year_per_capita,
        poultry = meat__poultry__00002734__food_available_for_consumption__0645pc__kilograms_per_year_per_capita) %>%
  pivot_longer(cols = other:poultry,
               names_to = "meat_type",
               values_to = "kg_y_pc") %>% 
  mutate(kg_y_pc= ifelse(is.na(kg_y_pc), 0, kg_y_pc)) %>% 
    mutate(meat_type= ifelse(is.na(meat_type), 0, meat_type))
```

```{r, production}
#meat production (mp) 
url<-'https://ourworldindata.org/grapher/global-meat-production-by-livestock-type.csv?v=1&csvType=full&useColumnShortNames=true'
OGmeat_production<-read_csv(url)

production<- OGmeat_production%>% rename(beef=meat__beef_and_buffalo__00001806__production__005510__tonnes,
poultry = meat__poultry__00001808__production__005510__tonnes) %>% 
  mutate(beef=ifelse(is.na(beef), 0, beef), poultry=ifelse(is.na(poultry), 0, poultry)) %>% 
  select(year,entity, beef, poultry) %>% 
  pivot_longer(cols =beef:poultry,
            names_to="meat_type") %>% 
 rename(tons=value)

```


```{r}
mp<- OGmeat_production %>% 
  rename(beef=meat__beef_and_buffalo__00001806__production__005510__tonnes,
        poultry=meat__poultry__00001808__production__005510__tonnes,
           game = meat__game__00001163__production__005510__tonnes,
         horse = meat__horse__00001097__production__005510__tonnes,
         camel = meat__camel__00001127__production__005510__tonnes,
         sheep_goat= meat__sheep_and_goat__00001807__production__005510__tonnes,
         pig = meat__pig__00001035__production__005510__tonnes) %>% 
  mutate(beef=ifelse(is.na(beef), 0, beef), 
         poultry=ifelse(is.na(poultry), 0, poultry),
         horse = ifelse(is.na(horse), 0, horse),
         pig = ifelse(is.na(pig), 0, pig),
         camel = ifelse(is.na(camel), 0, camel),
         beef = ifelse(is.na(beef), 0, beef),
         poultry = ifelse(is.na(poultry), 0, poultry),
         game = ifelse(is.na(game), 0, game),
         sheep_goat = ifelse(is.na(sheep_goat), 0, sheep_goat)) %>% 
pivot_longer(cols = game:poultry,
             names_to="meat_type") %>% 
  rename(tons=value) %>% filter (entity %in% c( carnivores, high_risk))

#new defs###################################################################
carnivores<-c('United States', 'Argentina', 'Australia')
high_risk<-c('Yemen', 'Afghanistan', 'Haiti')

nmp<-mp %>% filter(entity %in% c(carnivores, high_risk))
nmgdp<-mgdp %>% filter(entity %in% c(carnivores, high_risk))
nmpc<-mpc%>% filter(entity %in% c(carnivores, high_risk))%>% 
  mutate(country_cat = factor(entity, levels = c('Argentina', 
  'Australia', 'United States', 'Haiti', 'Afghanistan', 'Yemen'))) %>% mutate(meat_type=factor(meat_type,levels = c('beef', 'poultry', 'pig', 'sheep_goat', 'other')))

```

## The grim reality of the present 

Approximately **9 million people** die from hunger-related causes annually. 

~ 1 person dies / 4 seconds

~ Malnutrition is the leading cause of child mortality, contributing to roughly 50% of all child deaths 

~ Over 3 million children die from hunger annually.

## UN SDGs 

According to the UN, hunger is the leading cause of death in humans, but this is not purely a shortage issue. The Huffington Post states, “The world already produces enough food for everyone, but around **one third** of it is discarded or spoils in transport or storage before reaching consumers, according to the FAO. Additionally, High-income countries account for just *16%* of the world’s population and *33%* of its meat intake.” This begs the question- if we improved how we produced and distributed food (specifically meat), could we eliminate one of the leading causes of death in our species? 

In this presentation we will examine meat consumption and production patterns under the lens of the **2nd and 12th** United Nations Sustainable Development Goals **(Zero Hunger and Responsible Production & Consumption)** to see how they might be altered.

## Introduction to Sample Population

**Haiti, Yemen**, and **Afghanistan** are three of the seven countries of greatest concern for hunger risk according to *Action Against Hunger*. These countries will serve as our `famine high-risk` group. 

This group will contrast the `carnivore` group, which is made up of the **United States, Argentina**, and **Australia** - some of the biggest meat-consuming countries in the world (per capita).

## data 1

```{r}
ggplot(nmgdp,
       aes(x=year, y=gdp_per_capita, color=entity))+geom_point()+
  labs(title="gdp per capita for countries over the years") 
```

## data 2

```{r}
ggplot(nmp, aes(x = entity, y = tons,fill=meat_type)) +
    geom_col(position = "dodge")+labs(title='meat produced per capita')+
  scale_y_continuous(labels = scales::label_comma()) 
```

## data 3

```{r}
ggplot(nmpc, aes(x = entity, y = kg_y_pc,fill=meat_type)) +
    geom_col(position = "dodge")+labs(title='meat consumed per capita')
```



## Meat efficiency 
<u>Feed efficiency</u>, a critical measure in livestock production, determines how effectively animals convert feed into desired outputs like milk, eggs, or meat. 

Production Efficiency also correlates with the **environmental impact** these productions have.
According to a 2014 study, diets must feature less red meat, which consumes 11 times more water and results in five times more climate-warming emissions than chicken or pork.

## Chicken vs Beef

+ **Broiler Chickens:** Broiler chickens have an exceptional Feed Efficiency Ratio (FER), often around 1.5 to 2.0. 
    + This means for every 1.5 to 2 pounds of feed, the chicken produces 1 pound of meat. 

+ **Beef Steers:** Beef steers have a feed efficiency ratio of approximately 6 to 8. 
    + This lower efficiency is due to their longer growth periods and higher maintenance energy requirements compared to poultry and pigs.

## With these insights, the goal seems to be to shift consumption away from beef and towards poultry.
```{r}
ggplot(meat_percap, 
       aes(x=year, 
           y=meat__beef_and_buffalo__00002731__food_available_for_consumption__0645pc__kilograms_per_year_per_capita, 
           color=entity))+
  geom_point()+
  labs(title='Beef Consumption',
       x='Year', y='Beef (kg/year/per capita)')+
  scale_y_continuous(labels = scales::label_comma())
```

## Which is what we are already seeing! 

```{r}
ggplot(meat_percap,
       aes(x=year, 
           y=meat__poultry__00002734__food_available_for_consumption__0645pc__kilograms_per_year_per_capita, 
           color=entity))+
  geom_point()+
  labs(title='Poultry Consumption', x='Year', y='Poultry(kg/year/per capita)')+
  scale_y_continuous(labels = scales::label_comma()) 

```


## Production trends shows some promise ! 
```{r}
ggplot(production,
       aes(x=year, 
           y=tons, color=meat_type))+
  geom_point()+
  labs(title='Poultry + Beef Production', x='Year', y='Beef + Poultry(tonnes)')+
  scale_y_continuous(labels = scales::label_comma())
```

## Is it really worth it/doable? (ending)

+ "I don't think it's all that ambitious to eliminate hunger," said Jomo Sundaram, assistant director-general of the U.N.'s Food and Agriculture Organisation (FAO). That is because **incomes are rising** in much of the world, **transport to move food is improving**, and **new technologies** are keeping yields of many key crops on an upward trend, he said.

+ Despite a rapidly rising world population, *there are 216 million fewer hungry people on earth today than in 1990*, the FAO reported in May.

